\section{KR and Exercises: Configuration}

\subsection{Configuration Clock Speed Calculation}

\begin{example2}{Configuration Clock Speed}

\textbf{Question:} The table below is an excerpt from the Cyclone V datasheet. It lists the configuration file size in bits for the different device sizes (A2–A7) in the Cyclone V E series.

\begin{center}
\begin{tabular}{|l|c|c|}
\hline
\textbf{Variant} & \textbf{Member Code} & \textbf{Configuration .rbf Size (bits)} \\
\hline
\multirow{4}{*}{Cyclone V E$^{(97)}$} & A2 & 21,061,280 \\
\cline{2-3}
& A4 & 21,061,280 \\
\cline{2-3}
& A5 & 33,958,560 \\
\cline{2-3}
& A7 & 56,167,552 \\
\hline
\end{tabular}
\end{center}

Your design contains \textbf{one} Cyclone V E \textbf{A2} device. The \textbf{configuration time must not exceed 500 ms}. Your design will use \textbf{serial configuration mode}.

From the list, select the \textbf{lowest configuration clock speed} that still meets the time requirement. \underline{\hspace{2cm}} MHz.

\tcblower

\textbf{Explanation:}

Correct answer: \textbf{44 MHz}

Calculation:
\begin{itemize}
    \item Configuration file size: 21,061,280 bits
    \item Required time: 500 ms = 500,000 $\mu$s
    \item Clock speed = $\frac{\text{number of bits}}{\text{time in microseconds}}$ = $\frac{21,061,280}{500,000}$ = 42.12 MHz
    \item Lowest suitable clock speed: 44 MHz
\end{itemize}

Divide number of bits by required time in microseconds to get the clock speed in MHz.

\important{Problem type:} Configuration clock speed calculation

\textbf{Source:} Moodle Quiz Configuration, Question 1
\end{example2}

\subsection{Configuration Time Calculation}

\begin{example2}{Configuration Time}

\textbf{Question:} The table below is an excerpt from the Cyclone V datasheet. It lists the configuration file size for the different device sizes (A2–A7) in the Cyclone V E series.

\begin{center}
\begin{tabular}{|l|c|c|}
\hline
\textbf{Variant} & \textbf{Member Code} & \textbf{Configuration .rbf Size (bits)} \\
\hline
\multirow{4}{*}{Cyclone V E$^{(97)}$} & A2 & 21,061,280 \\
\cline{2-3}
& A4 & 21,061,280 \\
\cline{2-3}
& A5 & 33,958,560 \\
\cline{2-3}
& A7 & 56,167,552 \\
\hline
\end{tabular}
\end{center}

Your design contains \textbf{one Cyclone V A5} device. The \textbf{configuration clock is 40 MHz}.

How long will configuration take for \textbf{a single A5} device when using \textbf{serial mode}? \underline{\hspace{3cm}} us (microseconds)

\tcblower

\textbf{Explanation:}

Correct answer: \textbf{848964 $\mu$s} (microseconds)

Calculation:
\begin{itemize}
    \item Configuration file size: 33,958,560 bits
    \item Clock frequency: 40 MHz = 40 bits/$\mu$s (in serial mode)
    \item Configuration time = $\frac{\text{number of bits}}{\text{clock frequency}}$ = $\frac{33,958,560}{40}$ = 848,964 $\mu$s
\end{itemize}

\important{Problem type:} Configuration time calculation

\textbf{Source:} Moodle Quiz Configuration, Question 2
\end{example2}

\raggedcolumns
\columnbreak