\section{Recommender Systems}

\begin{definition}{Good Recommender System}\\
A good recommender system...
\begin{itemize}
    \item Recommends personalized and valuable items within the current context
    \item Recommends items from all different and possible areas of interest
    \item Does not recommend items which are already known
    \item Extends the areas of interest (serendipity effect)
\end{itemize}
\end{definition}

\subsection{Collaborative Filtering}

\begin{concept}{Collaborative Filtering}\\
Analysis of behaviour patterns of user groups. Recommends items to a user based on comparison of their behaviour against users with a "similar" behaviour.
\end{concept}

\begin{formula}{Cosine Similarity}\\
$$similarity = \cos(\theta) = \frac{A \cdot B}{||A|| \cdot ||B||} = \frac{\sum_{i=1}^{n} A_i B_i}{\sqrt{\sum_{i=1}^{n} A_i^2} \cdot \sqrt{\sum_{i=1}^{n} B_i^2}}$$
\end{formula}

\begin{concept}{Challenges in Recommender Systems}
\begin{itemize}
    \item Many items to recommend, but only a few can be recommended
    \item Mostly sparse data per user
    \item Interest measure can be very diverse (Explicit: ratings, Implicit: clicks)
    \item No data for new users
\end{itemize}
\end{concept}

\subsection{Types of Filtering}

\begin{definition}{Content-Based Collaborative Filtering}\\
Recommendations are based on the content or the attributes of the objects instead of the user behaviour and ratings. There are explicit attributes and the content of an object.
\end{definition}

\begin{definition}{Context-Aware Filtering}\\
Filtering based on a dynamic collection of factors that describe the status of a user (time, location, social factor).
\end{definition}

\subsection{User-Based Collaborative Filtering}

\begin{concept}{User-Based Collaborative Filtering}\\
Calculates the similarity between the users based on their rating behaviour.
\end{concept}

\begin{example2}{User Similarity Example}\\
Given user vectors:
\begin{itemize}
    \item $A = (4, 3, 0, 0, 5)^T$
    \item $B = (5, 0, 4, 0, 4)^T$
    \item $C = (4, 0, 5, 3, 4)^T$
\end{itemize}

Calculate similarity:
$$sim_{AB} = \frac{(4 \cdot 5) + (3 \cdot 0) + \cdots + (5 \cdot 4)}{\sqrt{16 + 9 + 25} \cdot \sqrt{25 + 16 + 16}} = 0.75$$

Recommend items with a high rating from these similar users:
$$\frac{0.75 \cdot 4 + 0.63 \cdot 5}{0.75 + 0.63}$$
\end{example2}

% NOTE: Add user-item rating matrix visualization

\subsection{Item-Based Collaborative Filtering}

\begin{concept}{Item-Based Collaborative Filtering}\\
Calculates the similarity between the items based on their rating.
\end{concept}

\begin{example2}{Item Similarity Example}\\
Given item vectors:
\begin{itemize}
    \item $A = (4, 5, 4, 0, 0)^T$
    \item $B = (3, 0, 0, 3, 4)^T$
\end{itemize}

$$sim_{AB} = \frac{(4 \cdot 3)}{\sqrt{16 + 25 + 16} \cdot \sqrt{9 + 9 + 16}} = 0.27$$

Recommends the most similar items with the highest ranking.
\end{example2}

% NOTE: Add item similarity matrix visualization